\sysfull

### 2018-08-06 paper/asplos19/coalescing.tex
Task coalescing recectioned
Get rid of redundancies in sections 4, 5 and 6

When a device's buffered energy is sufficient to run multiple consecutive tasks without a power failure, committing state after each task is an unnecessary overhead. \sys eliminates this overhead by coalescing a sequence of tasks and deferring the commit
operations for all tasks to the end of the sequence. In general, committing involves moving state manipulated by a task into its permanent location in non-volatile memory. In particular, \sys's commit procedure copies dirty pages of memory that a task updated between fast, volatile working memory and slower, non-volatile main memory; we defer the details of paging privatization and commit to Section~\ref{sec:memory_virtulaization}.

Coalescing has a double benefit. First, coalescing reduces the total number of relatively high-latency, high-energy non-volatile memory accesses because tasks primarily access variables in relatively low-latency, low-energy volatile memory. Second, coalescing eliminates the unnecessary instructions that implement task commit, which constitute a substantial run-time overhead, as our measurements in Section~\ref{sec:evaluation} show.

An effective coalescing strategy must be {\em aggressive} enough, attempting to coalesce a large number of tasks to amortize commit overhead. However, it must also be  {\em conservative} enough, attempting to coalesce only as many tasks as will execute to completion given a certain energy conditions, reducing the risk or long coalesced task re-execution penalty.

\begin{algorithm}[t]
	\caption{Coalescing}
	\label{algo:genCoalescing}
	\scriptsize
	\begin{algorithmic}[1]
        \State $B \leftarrow $ \Call{$f_\text{reboot}$}{$z$}
        \State $H \gets 0$ \Comment{\note{add comments for some lines to get rid of white spaces}}
        \While{$\texttt{true}$}       
	        \State $ i \gets 0$
	        %
	        \While{$ i <= B$}
		        \State \Call{execute\_task}{$T_\texttt{i}$}
		        \State $W \leftarrow $ \Call{$f_\text{weight}$}{$T_\text{i}$}
		        \State $i \gets i + W$
				\State $H \gets H + W$
	        \EndWhile
	        %
	        \State \Call{commit\_to\_fram}{\null}
	        \State $B \leftarrow $ \Call{$f_\text{commit}$}{$B$}
        \EndWhile
	\end{algorithmic}
\end{algorithm}

\subsection{Task Coalescing Strategies}
\label{subsec:coalescingStrategies}

Algorithm~\ref{algo:genCoalescing} shows the general structure of a coalescing strategy. In the algorithm, $B$ (target budget) is total number of static tasks that \sys will next attempt to coalesce. $T_\text{i}$ is the $i$th task executed since the last power failure. History, $H$, is the total number of tasks between the last executed task and the power failure that preceding it. $W_\text{i}$ is the {\em weight} of a task $T_\text{i}$. A task's weight $W$ is an arbitrary quantity associated with the task that represents its cost in time or energy to execute; different coalescing strategies may apply different weights to a task. For example, a uniform weighting scheme sets the weight of each task to one, making them all effectively equal in cost to the coalescing algorithm.  A different policy might use a weighting scheme that sets a task's weight equal to its run time relative to the task that executes for the shortest duration, which has unity weight.

Different coalescing strategies adhere mainly to the template in Algorithm~\ref{algo:genCoalescing}, varying in only a few {\em characteristic operations} that the algorithm leaves deliberately abstract. A strategy's reboot update function, $f_\text{reboot}$, updates $B$, the target budget, after a reboot. A strategy's weight lookup function, $f_\text{weight}$, returns the a task's weight.  A strategy's commit update function, $f_\text{commit}$, updates $B$ after a successful commit. The following sections detail several possible coalescing strategies that we developed for \sys, albeit not the only possible. In fact, \sys \emph{allows programmers to design their own strategy} to implement and link to \sys's task scheduler.

\subsubsection{Energy-Oblivious Coalescing}
\label{subsec:energyBlind}

A very simple {\em energy-oblivious} (EO) coalescing strategy treats all tasks as having unity weight and varies the size of a coalesced task linearly. In such a scheme, the number of tasks to coalesce, $B$, increases by a constant $x$ when a coalesced sequence of tasks commits, and decreases by the same constant $x$ when a coalesced sequence of tasks fails to complete, i.e., after a power failure. The characteristic operations of the energy-oblivious strategy are
%
\begin{equation}
	\begin{split}
		f_\text{reboot}(B) & = B - x, \\
		f_\text{weight}(T_\text{i}) & =  1, \\
		f_\text{commit}(B) & = B + x.
	\end{split}
	\label{eq:eo}
\end{equation}
%
The energy-oblivious strategy reacts slowly to the variation in the energy required to execute different tasks and to the variation in the effective quantum of energy available to the device. With the successful commit of each coalesced task sequence, the coalescing target increases. Eventually, the target may be too high and a coalesced task composed of only fewer tasks will commit without interruption by a power failure. The strategy then linearly decreases the coalescing target, eventually reaching a coalescing target that allows completion. A key limitation of this algorithm lies in the equality of the target decrease in $f_\text{reboot}$ and the increase in $f_\text{commit}$. Let us assume $x=1$. If, after $k$ successful commits the target must decrease to its original value $B-k$ due to an energy drop, this strategy requires $k$ successive power failures.

\subsubsection{Energy-Guided Coalescing}
\label{subsec:energyAware}

\begin{figure*}
    \includegraphics[width=\linewidth]{figures/hg-coal-horiz.pdf}
    \caption{HG coalescing sample execution across two power cycles. \textcolor{red}{explain the meaning of the letters and the logical connection between the left and the right (or cycle 1 and 2)}}
    \label{fig:hg-coal}
\end{figure*}

The \emph{Energy-Guided} (EG) coalescing strategy adapts its coalescing target more quickly, addressing a key limitation of the EO, to adhere to changes in energy conditions. It uses its recent execution history as  mean to measure energy availability and alters its target accordingly. By relying on the history of execution EG eliminates the problem of frequent power failures a single coalesced task. In fact EG can reduce its target to one from any coalesced target in at most two successive power failures. For instance, if the execution is interrupted after finishing a single task the history will be one ($H = 1$) and the target will be set to one also. While a history of $H$ suggests that there is sufficient energy to coalesce and successfully commit $H$ tasks (i.e., to set $B = H$). EG conservatively sets $B$ equal to half of $H$ moderating the risk of a very long coalesced task. We experimented with a range of fraction values, settling on one half as the default for EG, because one half had the highest performance. The EG algorithm is characterized by the following functions
%
\begin{equation}
	\begin{split}
		 f_\text{reboot}(H) & = \lceil(H / 2)\rceil,\\
		 f_\text{weight}(T_\text{i}) & = 1, \\
		 f_\text{commit}(B) & = \lceil(B / 2)\rceil.
	\end{split}
	\label{eq:eg}
\end{equation}
%
Fig.~\ref{fig:hg-coal} illustrates the operation of the EG. The figure shows two power cycles following the $i$th, the latter having an
execution history $H^{(i)} = 4$. Based on that, $B$ is initially set to two and then decreases to one. Once the value of $B$ reaches one, EG is, in effect, expecting a power failure, justifying the conservative approach. After the power failure, EG uses the most recent history $H^{(i+1)} = 8$ to set $B = 4$ and continue execution.

\subsubsection{Weighted Energy-Guided Coalescing}
\label{subsec:energyTaskAware}

The \emph{Weighted Energy-Guided} (WEG) coalescing strategy accounts for the different energy and time cost of a program's tasks when setting the coalescing target. Each different task in the program consumes a different amount of energy to execute to completion. The EO and EG coalescing strategies assume each task has uniform energy cost: for these strategies, $B$ simply corresponds to a target {\em count} of tasks coalesced, regardless of the individual cost of each task. However, if one task executes for ten seconds and another executes for one second, counting tasks misjudges the amount of {\em work} in the coalesced tasks.

Instead, WEG associates a non-unity weight with each task in the program. WEG tracks the sum of the weights of tasks in a coalesced sequence of tasks. When the sum of weights reaches $B$, the target, WEG commits the coalesced task. WEG differs from EO and HG by explicitly accounting for the variation in energy cost of a program's tasks, eliminating the assumption that all tasks do the same work, which is a key limitation. WEG is characterized the same way as HG (equation (\ref{eq:eg})), except that $f_\text{weight}(T_\text{i}) = W_\text{i}$.

The effectiveness of the WEG hinges on correctly identifying the weight of each task, which WEG assumes is {\em statically} available. Profiling the time and energy cost of tasks in a program is a difficult, orthogonal problem~\cite{cleancut_2018,baghsorkhi_cgo_2018}. WEG could use the result of an arbitrarily sophisticated profiling procedure. To produce a concrete result in this paper, we give WEG access to a simple profile of task run time collected offline using a single fixed input. WEG stores the profile in a lookup table that maps a task's identifier to its weight, making the information available to \sys's scheduler at run time.

>>>>

When a device's buffered energy is sufficient to run multiple consecutive tasks without a power failure, committing state after each task is an unnecessary overhead. \sys eliminates this overhead by coalescing a sequence of tasks and deferring the commit
operations for all tasks to the end of the sequence. In general, committing involves moving state manipulated by a task into its permanent location in non-volatile memory. In particular, \sys's commit procedure copies dirty pages of memory that a task updated between fast, volatile working memory and slower, non-volatile main memory. We defer the details of paging privatization and commit to Section~\ref{sec:memory_virtulaization}.
An effective coalescing strategy must be {\em aggressive} enough, attempting to coalesce a large number of tasks to amortize commit overhead.
However, it must also be  {\em conservative} enough, attempting to coalesce only as many tasks as will execute to completion given certain energy conditions, reducing the risk or long coalesced task re-execution penalty.

\subsubsection{Generic Design of Task Coalescing Strategies}
\label{subsec:coalescingGeneral}

% Coalescing has a double benefit. First, coalescing reduces the total number of relatively high-latency, high-energy non-volatile memory accesses because tasks primarily access variables in relatively low-latency, low-energy volatile memory. Second, coalescing eliminates the unnecessary instructions that implement task commit, which constitute a substantial run-time overhead, as our measurements in Section~\ref{sec:evaluation} show.

\begin{algorithm}[t]
	\caption{Coalescing}
	\label{algo:genCoalescing}
	\scriptsize
	\begin{algorithmic}[1]
        \State $C \leftarrow $ \Call{$f_\text{reboot}$}{$z$}
        \State $H \gets 0$ \Comment{\note{add comments for some lines to get rid of white spaces}}
        \While{$\texttt{true}$}       
	        \State $ i \gets 0$
	        %
	        \While{$ i <= C$}
		        \State \Call{execute\_task}{$T_\texttt{i}$}
		        \State $W \leftarrow $ \Call{$f_\text{weight}$}{$T_\text{i}$}
		        \State $i \gets i + W$
				\State $H \gets H + W$
	        \EndWhile
	        %
	        \State \Call{commit\_to\_fram}{\null}
	        \State $C \leftarrow $ \Call{$f_\text{commit}$}{$C$}
        \EndWhile
	\end{algorithmic}
\end{algorithm}

Algorithm~\ref{algo:genCoalescing} shows the general structure of a coalescing strategy. In the algorithm, $C$ (coalescing target) is total number of static tasks that \sys will next attempt to coalesce. $T_\text{i}$ is the $i$th task executed since the last power failure. $H$ (history) is the total number of tasks executed since the last power failure. $W_\text{i}$ is the {\em weight} of a task. A task's weight is an arbitrary quantity associated with the task that represents its cost in time or energy to execute. Different coalescing strategies may apply different weights to a task,
e.g., $W_\text{i} = 1\ \forall i$, or $W_\text{i} = \alpha E(T_\text{i})$ where $E(T_\text{i})$ is the average execution time of $T_\text{i}$ and $\alpha$ is a constant.
% For example, a uniform weighting scheme sets the weight of each task to one, making them all effectively equal in cost to the coalescing algorithm.  A different policy might use a weighting scheme that sets a task's weight equal to its run time relative to the task that executes for the shortest duration, which has unity weight.

\subsubsection{Example Task Coalescing Strategies}
\label{subsec:coalescingStrategies}

Different coalescing strategies adhere mainly to the template in Algorithm~\ref{algo:genCoalescing}, varying in only a few {\em characteristic operations} that the algorithm leaves deliberately abstract. A strategy's reboot update function, $f_\text{reboot}$, updates $C$, the coalescing target, after a reboot. A strategy's weight lookup function, $f_\text{weight}$, returns a task's weight.  A strategy's commit update function, $f_\text{commit}$, updates $C$ after a successful commit. The following paragraphs detail several possible coalescing strategies that we developed for \sys, albeit not the only possible. In fact, \sys \emph{allows programmers to design their own strategy} to implement and link to \sys's task scheduler.

\paragraph{Energy-Oblivious Coalescing.}
\label{subsec:energyBlind}

A very simple {\em energy-oblivious} (EO) coalescing strategy treats all tasks as having unity weight and varies the size of a coalesced task linearly. In such a scheme, the number of tasks to coalesce, $C$, increases by a constant $x$ when a coalesced sequence of tasks commits, and decreases by the same constant $x$ when a coalesced sequence of tasks fails to complete, i.e., after a power failure. The characteristic operations of EO are
%
\begin{equation}
	\begin{split}
		f_\text{reboot}(C) & = C - x, \\
		f_\text{weight}(T_\text{i}) & =  1, \\
		f_\text{commit}(C) & = C + x.
	\end{split}
	\label{eq:eo}
\end{equation}
%
The energy-oblivious strategy reacts slowly to the variation in the energy required to execute different tasks and to the variation in the effective quantum of energy available to the device. With the successful commit of each coalesced task sequence, the coalescing target increases. Eventually, the target may be too high and only a coalesced task composed of a few units will commit without interruption by a power failure. The strategy then linearly decreases $C$, eventually reaching a value that allows completion. A key limitation of this algorithm lies in the equality of the target decrease in $f_\text{reboot}$ and the increase in $f_\text{commit}$. Let us assume $x=1$. If, after $k$ successful commits, the target must decrease to its original value $C-k$ due to an energy drop, EO requires $k$ successive power failures.

\paragraph{Energy-Guided Coalescing.}
\label{subsec:energyAware}

\begin{figure*}
    \includegraphics[width=\linewidth]{figures/hg-coal-horiz.pdf}
    \caption{EG coalescing sample execution across two power cycles. C: coalescing target, H: history.}
    \label{fig:hg-coal}
\end{figure*}

The \emph{Energy-Guided} (EG) coalescing strategy adapts its coalescing target more quickly, addressing a key limitation of the EO, to adhere to changes in energy conditions. It uses its recent execution history H to estimate energy availability and alters its target accordingly. The EG algorithm is characterized by the following functions
%
\begin{equation}
	\begin{split}
		 f_\text{reboot}(H) & = \lceil \rho H \rceil,\\
		 f_\text{weight}(T_\text{i}) & = 1, \\
		 f_\text{commit}(C) & = \lceil \gamma C \rceil,
	\end{split}
	\label{eq:eg}
\end{equation}
%
where $\rho$ and $\gamma$ belong to $[0, 1]$.
By relying on the history of execution EG eliminates the problem of frequent power failures on a single coalesced task.
In fact, at any reboot, EG will {\em conservatively} try to coalesce only a fraction ($\rho$) of tasks that had successfully completed during the last power cycle.
% In fact EG can reduce its target to one from any coalesced target in at most two successive power failures. For instance, if the execution is interrupted after finishing a single task the history will be one ($H = 1$) and the target will be set to one also. While a history of $H$ suggests that there is sufficient energy to coalesce and successfully commit $H$ tasks (i.e., to set $C = H$). EG conservatively sets $C$ equal to half of $H$ moderating the risk of a very long coalesced task. We experimented with a range of fraction values, settling on one half as the default for EG, because one half had the highest performance.
Figure~\ref{fig:hg-coal} illustrates a snapshot of the operation of EG, with $\rho = \gamma = 0.5$. The figure shows two power cycles following the $i$th, the latter having an
execution history $H^{(i)} = 4$. Based on that, $C$ is initially set to two and then decreases to one. Once the value of $C$ reaches one, EG is, in effect, expecting a power failure, justifying the conservative approach. After the power failure, EG uses the most recent history $H^{(i+1)} = 8$ to set $C = 4$ and continue execution.

\paragraph{Weighted Energy-Guided Coalescing.}
\label{subsec:energyTaskAware}

The \emph{Weighted Energy-Guided} (WEG) coalescing strategy accounts for non-uniform energy and time costs of a program's tasks when setting $C$. Each different task in the program consumes a different amount of energy to execute to completion. The EO and EG coalescing strategies assume that each task has the same cost: for these strategies, $C$ simply corresponds to a target {\em count} of tasks to coalesce, regardless of the individual cost of each task. However, if one task executes for ten seconds and another executes for one second, counting tasks misjudges the amount of {\em work} in the coalesced tasks (and the history).
%
Instead, WEG associates a non-unity weight with each task in the program, and tracks the sum of the weights of tasks in a coalesced sequence. When the sum of weights reaches $C$, the target, WEG commits the coalesced task.
% WEG differs from EO and HG by explicitly accounting for the variation in energy cost of a program's tasks, eliminating the assumption that all tasks do the same work, which is a key limitation.
WEG is characterized the same way as EG (equation (\ref{eq:eg})), except that $f_\text{weight}(T_\text{i}) = W_\text{i}$.
%
% The effectiveness of the WEG hinges on correctly identifying the weight of each task, which WEG assumes is {\em statically} available. Profiling the time and energy cost of tasks in a program is a difficult, orthogonal problem~\cite{cleancut_2018,baghsorkhi_cgo_2018}. WEG could use the result of an arbitrarily sophisticated profiling procedure. To produce a concrete result in this paper, we give WEG access to a simple profile of task run time collected offline using a single fixed input. WEG stores the profile in a lookup table that maps a task's identifier to its weight, making the information available to \sys's scheduler at run time.

### 2018-08-06 paper/asplos19/paging.tex
Get rid of redundancies in sections 4, 5 and 6

%\note{additional motivation for coala memory virtualization] we should emphasize that Coala programming model removes the burden of identifying the write-after-read dependencies from the programmer  as compared to Chain. And it does not require a specific compiler that cannot consider a runtime dependent behavior like Interrupts service handling to guarantee memory protection, as in Alpaca.}

\sys's {\em virtual memory manager} is a core enabler of \sys task adaptation mechanism. Its overview is given in Figure~\ref{figure:mem-man}. VMM abstracts the physical address space of the non-volatile memory (FRAM) and divides it into \texttt{\underline{private}} and \texttt{\underline{shadow}} (underlined locations are non-volatile) buffers. Each buffer is divided into pages. The VMM prohibits applications from directly access these buffers. Instead, it redirects any request to the \texttt{working} buffer located in the volatile memory (SRAM), which has a lower latency and energy cost to access than FRAM. It populates \texttt{working} buffer with the privatized pages from non-volatile memory. When a coalesced task completes, the VMM ensures that updated pages are committed to main memory and visible to subsequently executed tasks. If the power interrupts a task, the VMM ensures the consistency of the data by relying on its \texttt{\underline{private}} buffer. The VMM ensures consistency even if a task or a sequence of tasks include WAR dependences that could compromise consistency~\cite{ratchet,dino}. On a commit, the VMM atomically copies, through two-phase commit, all ``dirty'' pages back into their location in the non-volatile memory. SRAM has a limited size and there is an extended working memory containing {\em shadow pages} in the larger FRAM. A shadow page stores the data from an updated pages that need to be committed back to non-volatile memory, but that exceeded the size of the volatile working memory. The VMM moves pages of data efficiently using hardware accelerated DMA support.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/mem-man.pdf}
    \caption{Interaction between task, memory manager and memory buffers upon \texttt{RP} and \texttt{WP}, and its order of occurrence; steps 2, 3, 4 and 5 are conditional. Tasks only have direct access to the \texttt{working} buffer.}
    \label{figure:mem-man}
\end{figure}

@@

A task must access protected non-volatile variables through \sys's restricted memory interface. The interface includes \texttt{RP}$(v)$, to read the value of variable $v$, and \texttt{WP}$(v)$ to assign a value to variable $v$. The implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt WP}'s implementation is similar except that {\tt WP} marks the accessed page as dirty. {\tt RP} and {\tt WP} operations translate a variable's physical address in non-volatile memory into a \emph{virtual address} in \texttt{working} buffer in volatile memory. In \sys, a virtual address is composed of a \emph{page tag} that identifies the page and a \emph{page offset} that identifies a byte.

@@

When accessing a protected variable with \texttt{RP} or \texttt{WP}, the memory manager first searched the variable's page in the \texttt{working} buffer (Figure~\ref{figure:mem-man}, Step 1). If the page is not found there, a page fault is incurred and a new page needs to
be swapped in. If the \texttt{working} buffer is full, a page fault on a memory access requires VMM to swap out one of the pages in the \texttt{working} buffer (a \emph{victim page}) preserving updates made to that page. While we opted for the simplicity of FIFO, any other replacement policy, such as Least Recently Used (LRU) could work in its place. If a task modified any byte in the victim page (i.e., using \texttt{WP}), then the page is dirty and its changes has to be persisted to non-volatile memory (Figure~\ref{figure:mem-man}, Step 2). If the accessed page was previously modified and swapped out since the last power failure, the most recent version of the page is in {\tt \underline{shadow}} (Step 3), otherwise it has to be retrieved from {\tt \underline{private}} (Step 4). Finally, the page is copied to the volatile buffer (Step 5).

@@

When the last task in a sequence of coalesced tasks completes, \sys must commit \emph{all} dirty pages in the \texttt{working} buffer and in the \texttt{\underline{shadow}} buffer, copying them back to their original locations in the main memory \texttt{\underline{private}}.

To make the commit atomic, \sys commits the pages in two phases, implemented by Algorithm~\ref{algo:commit}. The first phase copies dirty pages from the \texttt{working} buffer to the non-volatile \texttt{\underline{shadow}} buffer (Line 4). The second phase commits pages from \texttt{\underline{shadow}} to \texttt{\underline{private}} (Line 12).  If power fails during the first phase, the whole commit is aborted, and the execution restarts from the most recently committed point (i.e., from the beginning of a coalesced task). If power fails in the second
phase, the commit process safely resumes on reboot.  The second phase of commit depends on some run-time metadata. The \texttt{\underline{committing}} bit indicates that a commit is in progress and is set before the first page is committed (Line 9) and cleared after the last page is committed (Line 16).  The \texttt{\underline{shadowCount}} records the number of dirty shadow pages to commit and the VMM clears the counter when commit completes (Line 14). The \texttt{\underline{commitIndex}} indexes the next page to be committed (Line 11) and the VMM clears the index at the end of the phase (Line 15).

\sys's commit is efficient because it maintains an index dirty pages instead of iterating through all potentially dirty pages to check a dirty bit during commit. \sys's implementation is also efficient because the VMM does not copy from \texttt{\underline{shadow}} to \texttt{\underline{private}}, instead swapping pointers in an indirection table that maintains the pages as a double buffer.  Section~\ref{sec:implementation} describes double buffering in more detail.

\textbf{Memory Consistency.} \sys's paging mechanism ensures that a task only ever executes using consistent protected data. During task execution, modifications to protected data do not affect the \texttt{\underline{private}} buffer, because a task reads and writes the volatile \texttt{working} buffer only, and modified pages are kept in \texttt{\underline{shadow}} until commit. A power failure erases the contents of the \texttt{working} buffer, preventing a re-executing task from observing updates from a previous execution attempt. Clearing \texttt{\underline{shadowCount}} as part of the second phase commit (Line 14), ensures that all accesses to protected variables in subsequent tasks correctly access their original memory locations in \texttt{\underline{private}}.

@@

\sys asks the programmer to use its {\tt RP} and {\tt WP} paging API methods on every access to a protected variable (cf. Section~\ref{sec:coala_api}). These API invocations present a risk of high overhead because there is a dynamic check on every read and write. Despite the risk of per-access overhead, \sys's dynamic memory protection scheme brings several benefits over a static approach (i.e.~\cite{alpaca}). First, the limitations of static analysis preclude some uses of pointers due to potential pointer aliasing. For example, in the presence of arbitrary pointer operation, a function call using a function pointer, or an interrupt within a task, the system cannot statically analyze the memory behavior. Second, a static approach cannot handle task coalescing, because a protected variable's lifetime, i.e.,from first use to commit, is unknown at compile time. \sys's dynamic, per-access instrumentation supports arbitrary use of pointers and enables task coalescing.

>>>>

%\note{additional motivation for coala memory virtualization] we should emphasize that Coala programming model removes the burden of identifying the write-after-read dependencies from the programmer  as compared to Chain. And it does not require a specific compiler that cannot consider a runtime dependent behavior like Interrupts service handling to guarantee memory protection, as in Alpaca.}
%
\sys's {\em virtual memory manager} (VMM) is a core enabler of \sys task adaptation mechanism.
Its overview is given in Figure~\ref{figure:mem-man}.
VMM abstracts the physical address space of the non-volatile memory (FRAM) and divides it into \texttt{\underline{private}} and \texttt{\underline{shadow}} (underlined locations are non-volatile) buffers, and each buffer is divided into pages.
\texttt{\underline{private}} holds the consistent version of pages after each commit (and on reboot), while \texttt{\underline{shadow}} serves as support during the execution of coalesced tasks.
The VMM prohibits applications from directly accessing these buffers.
Instead, it redirects any request to the \texttt{working} buffer located in the volatile memory (SRAM), which has a lower latency and energy cost to access than FRAM.
It populates the \texttt{working} buffer with the privatized pages from non-volatile memory requested by tasks.
When a coalesced task completes, the VMM ensures that {\em dirty} (i.e., modified) pages are committed to FRAM and visible to subsequent tasks.
If a power failure interrupts a task, all temporary modifications to pages in \texttt{working} (and \texttt{\underline{shadow}}) occurred after the last commit are discarded in order to keep data consistency.
% If the power interrupts a task, the VMM ensures the consistency of the data by relying on its \texttt{\underline{private}} buffer. The VMM ensures consistency even if a task or a sequence of tasks include WAR dependences that could compromise consistency~\cite{ratchet,dino}.
On a commit, the VMM atomically copies, through a two-phase commit, all dirty pages back into their location in the non-volatile memory.
% SRAM has a limited size and there is an extended working memory containing {\em shadow pages} in the larger FRAM. A shadow page stores the data from an updated page that needs to be committed back to non-volatile memory, but that exceeded the size of the volatile working memory.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figures/mem-man.pdf}
    \caption{Interaction between task, memory manager and memory buffers upon \texttt{RP} and \texttt{WP}, and order of occurrence. Steps 2, 3, 4 and 5 are conditional.}
    \label{figure:mem-man}
\end{figure}

@@

A task must access protected non-volatile variables through \sys's restricted memory interface.
The interface includes \texttt{RP}$(v)$, to read the value of variable $v$, and \texttt{WP}$(v)$ to assign a value to variable $v$.
The implementation of {\tt RP} is shown in Algorithm~\ref{algo:rwar}, and {\tt WP}'s implementation is similar except that accessed page are marked as dirty.
{\tt RP} and {\tt WP} operations translate a variable's physical address in non-volatile memory into a \emph{virtual address} in \texttt{working} in volatile memory.
In \sys, a virtual address is composed of a \emph{page tag} that identifies the page and a \emph{page offset} that identifies a byte.

@@

When accessing a protected variable with \texttt{RP} or \texttt{WP}, the memory manager first searches the variable's page in the \texttt{working} buffer (Figure~\ref{figure:mem-man}, Step 1).
If the page is not found there, a page fault is incurred and a new page needs to be swapped in.
If the \texttt{working} buffer is full, a page fault on a memory access requires the VMM to swap out one of the pages in the \texttt{working} buffer (a \emph{victim page}) preserving updates made to that page.
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then the page is dirty and its changes have to be persisted to non-volatile memory (Figure~\ref{figure:mem-man}, Step 2). If the accessed page was previously modified and swapped out since the last power failure, the most recent version of the page is in {\tt \underline{shadow}} (Step 3), otherwise it has to be retrieved from {\tt \underline{private}} (Step 4). Finally, the page is copied to the volatile buffer (Step 5).

@@

When the last task in a sequence of coalesced tasks completes, \sys must commit \emph{all} dirty pages in \texttt{working} and \texttt{\underline{shadow}}, copying them back to their original locations in \texttt{\underline{private}}.

To make the commit atomic, \sys commits dirty pages in two phases, as shown in Algorithm~\ref{algo:commit}. The first phase copies dirty pages from \texttt{working} to the non-volatile \texttt{\underline{shadow}} (Line 4). The second phase commits pages from \texttt{\underline{shadow}} to \texttt{\underline{private}} (Line 12).  If power fails during the first phase, the whole commit is aborted, and the execution restarts from the most recently committed point (i.e., from the beginning of a coalesced task). If power fails in the second
phase, the commit process safely resumes on reboot.  The second phase depends on some run-time metadata. The \texttt{\underline{committing}} bit indicates that a commit is in progress and is set before the first page is committed (Line 9) and cleared after the last page is committed (Line 16).  The \texttt{\underline{shadowCount}} records the number of dirty shadow pages to commit and the VMM clears the counter when commit completes (Line 14). The \texttt{\underline{commitIndex}} indexes the next page to be committed (Line 11) and the VMM clears the index at the end of the phase (Line 15).

\sys's commit is efficient because it maintains an index of dirty pages instead of iterating through all potentially dirty pages to check their state.
Another source of efficiency lies in the second phase of the commit:
% \sys's implementation is also efficient because
the VMM does not copy pages from \texttt{\underline{shadow}} to \texttt{\underline{private}}, instead it swaps pointers in an indirection table that maintains the pages as a double buffer.

\paragraph{Memory Consistency.}
\sys's paging mechanism ensures that a task only ever executes using consistent protected data. During task execution, modifications to protected data do not affect the \texttt{\underline{private}} buffer, because a task reads and writes the volatile \texttt{working} buffer only, and modified pages are kept in \texttt{\underline{shadow}} until commit. A power failure erases the contents of the \texttt{working} buffer, preventing a re-executing task from observing updates from a previous execution attempt. Clearing \texttt{\underline{shadowCount}} as part of the second phase commit (Line 14) ensures that all accesses to protected variables in subsequent tasks correctly access their consistent memory locations in \texttt{\underline{private}}.

@@

\sys asks the programmer to use its {\tt RP} and {\tt WP} API methods on every access to a protected variable (cf. Section~\ref{sec:coala_api}). These API invocations present a risk of high overhead because there is a dynamic check on every read and write. Despite the risk of per-access overhead, \sys's dynamic memory protection scheme brings several benefits over a static approach (i.e.,~\cite{alpaca}). First, the limitations of static analysis preclude some uses of pointers due to potential pointer aliasing. For example, in the presence of arbitrary pointer operations, a function call using a function pointer, or an interrupt within a task, the system cannot statically analyze the memory behavior. Second, a static approach \textit{cannot handle task coalescing}, because a protected variable's lifetime, i.e., from first use to commit, is unknown at compile time. \sys's dynamic, per-access instrumentation supports arbitrary use of pointers and enables task coalescing.
